---
title: ""
subtitle: ""
author: 
  - Haowei Fan
  - Fangning Zhang
thanks: "Code and data are available at: "
date: today
date-format: long
abstract: ""
format: pdf
toc-depth: 2    
number-sections: true
toc: true           
toc-title: "Table of Contents" 
bibliography: references.bib
---

```{r}
#| echo: false
#| include: false
#| warning: false
#| message: false
#### Workspace setup ####
set.seed(912)
# List of necessary packages
packages <- c("ggplot2", "MASS", "pscl", "lmtest","GGally", "knitr","glmmTMB")
# Install missing packages
missing_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
install.packages(missing_packages)
# Load packages
lapply(packages, library, character.only = TRUE)

#### Read data ####
load(here::here("portugal.RData"))
```






# Methodology






# Result
@tbl-varibale presents the statistical summary of all important variables. It is shown that the distribution of The Number of Children is right-skewed, with a median and mean of 2. The interquartile range (1 to 3) shows that half of the population falls within this range, while the maximum value of 17 suggests high variability. The majority of individuals marry between the ages of 18 and 25, with the highest concentration in the 22–25 age group (1,468 individuals), followed by those aged 20–22 (1,126 individuals). Regarding the duration of marriage, the range extends from 0 months to 431 months. The median marriage duration is 148 months, with a mean of 155.2 months. However, the slightly higher mean contributes to a mild right-skewed distribution. In terms of literacy, the vast majority of the population is literate (4,567 individuals), whereas only 581 individuals are illiterate.

```{r}
#| warning: false
#| message: false
#| include: false
summary(portugal)
```
```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-varibale
#| tbl-cap: "Summary of Key Variables"
data_selec <- dplyr::select(portugal, children, monthsSinceM, ageMarried, literacy)
summary_df <- data.frame(
  "Number of Children" = c("Min: 0", "1st Qu: 1", "Median: 2", "Mean: 2", "3rd Qu: 3", "Max: 17", " "),
  "Months Since Marriage" = c("Min: 0.0", "1st Qu: 73.0", "Median: 148.0", "Mean: 155.2", "3rd Qu: 230.0", "Max: 431.0", " "),
  "Count by Age at Marriage" = c("At 0-15: 52", "At 15-18: 452", "At 18-20: 910", "At 20-22: 1126", "At 22-25: 1468", "At 25-30: 923", "At over 30: 217"),
  "Count by Literate or Illiterate" = c("Literate: 4567", "Illiterate: 581", " ", " ", " ", " ", " ")
  ,check.names = FALSE
)
kable(summary_df)
```


Recall our research objective: To explore the impact of literacy and age at marriage on family size (i.e., the number of children). We initially considered using Poisson regression because the number of children, as the response variable, is a non-negative integer. We included "Age at Marriage" and "Literate or Illiterate" as our predictor variables. Additionally, in @xxxx’s study, we have already found that the duration of marriage has a significant effect on the number of children. To control for the influence of different marriage durations on fertility, ensuring that the model reflects the independent effects of literacy and age at marriage, we included "Months of Marriage" as an offset to standardize the exposure period (i.e., the duration of marriage) across observations.

It is important to note that, since the link function of the Poisson distribution is log, directly taking log(0) when constructing the offset would result in negative infinity, which could affect the model's stability and accuracy. Therefore, when constructing the offset, we ensured that the number of months of marriage for each observation was at least 1 (even if the original data contained 0s, this adjustment prevents the log(0) issue). Additionally, we divided the months by 12 to convert the time scale to years, making the model's time representation more reasonable. By incorporating the offset, the original response variable (the number of children) is essentially transformed into a rate-based metric in the model. This means that we are now estimating the fertility rate per unit of time (e.g., per year) rather than the total number of children. With the logarithmic link function, this rate is expressed in logarithmic form, allowing the model’s results to be interpreted as the logarithm of the average number of children per year after marriage.

Furthermore, based on @tbl-variable, we observed that the majority of our sample got married between the ages of 22 and 25. Therefore, we set the base level for the "Age at Marriage" variable to the "22 to 25 years" group. This decision allows us to interpret the effects of other age-at-marriage categories relative to the most common marriage age group in our sample.

Thus, our model is:

\begin{align*}
\log(E[\text{Children}_i]) =& \beta_0 + \beta_1 \text{M}_{0to15}(i) + \beta_2 \text{M}_{15to18}(i) + \\ &\beta_3 \text{M}_{18to20}(i) +  \beta_4 \text{M}_{20to22}(i) + \beta_5 \text{M}_{25to30}(i) + \\ 
&\beta_6 \text{M}_{30toInf}(i) + \beta_8 \text{Literacy}_{i} + \\ &\log(\text{Months}_{i}) + \epsilon_i
\end{align*}
\begin{align*}
    \textbf{Where:} &\\
    \text{Children}_i & : \text{ The total number of children born to the } i\text{th individual.} \\
    M_{a\text{to} b}(i) & : \text{ Binary indicator variable, taking the value of } 1 \text{ if the } i\text{th individual } \\ 
    & \quad \text{entered into marriage between ages } a \text{ and } b, \text{ and } 0 \text{ otherwise.} \\
    \text{Literacy}_i & : \text{ Binary indicator variable, equal to } 1 \text{ if the } i\text{th individual is illiterate,} \\
    & \quad \text{and } 0 \text{ otherwise.} \\
    \text{Months}_i & : \text{ The total duration of marriage (in months) for the } i\text{th individual.} \\
    \beta_0, \beta_1, \dots, \beta_8 & : \text{ Coefficients estimated in the regression model.} \\
    \epsilon_i & : \text{ Stochastic error term capturing unobserved heterogeneity and random fluctuations.}
\end{align*}



```{r}
portugal$ageMarried = relevel(portugal$ageMarried, "22to25")
possi_model <- glm(children ~ ageMarried + literacy , 
                 family = poisson(link = "log"), 
                 data = portugal, 
                 offset = log(pmax(1, portugal$monthsSinceM)/12))
```

After determining the initial model, we examined its assumptions. We first observed the QQ plot (@fig-qqplot) and found that most of the residuals closely follow the reference line, indicating that the deviance residuals are approximately normally distributed. This suggests that the Poisson model is a reasonable fit for the data and confirms that the model satisfies the assumption that the log-mean of the response variable is linearly related to the predictor variables. Then, we calculated the dispersion parameter $\hat{\phi}$:

\begin{align*}
\hat{\phi} = \frac{\sum (\text{Pearson residuals})^2}{\text{Degrees of Freedom for Pearson Residuals}} = \frac{5600.133}{5140} = 1.08952 \approx 1
\end{align*}

$\hat{\phi}\approx 1$ indicates that the covariates in our model explain most of the variation, which suggests that the data do not exhibit significant overdispersion, and thus there is no need to switch the model to a negative binomial. Additionally, the assumptions that the observations are independent and that the response variable consists of positive integers are straightforward.

```{r, fig.pos="H"}
#| warning: false
#| echo: false
#| message: false
#| label: fig-qqplot
#| fig-cap: "QQ-Plot of The Final Model"

residuals_dev <- residuals(possi_model, type = "deviance")
qqnorm(residuals_dev, main = "")
qqline(residuals_dev, col = "red", lwd = 2) 
```
```{r}
#| warning: false
#| echo: false
#| message: false
#| include: false
overdispersion_ratio <- possi_model$deviance / possi_model$df.residual
possi_model$deviance
possi_model$df.residual
print(overdispersion_ratio)
```



@fig-rr illustrates the relative risk between different variable groups and their reference group. Compared to individuals who married between ages 22-25 (reference group), those who married at 0-15 or above 30 may have a slightly higher number of children (relative risk), but due to their wide confidence intervals covering 1, the reliability of this estimate is low. In contrast, those who married between 15-18 and 18-20 are more likely to have more children than reference group, as their confidence intervals are relatively narrow and do not include 1, indicating a stronger association. For individuals who married between 20-22 or 25-30, the number of children appears to remain nearly unchanged compared to the reference group, but since their confidence intervals include 1, no definitive conclusion can be drawn. For literacy, compared to the reference group (literacy), the illiteracy group has more children, and the confidence interval is very narrow, implying that there is a high probability that the illiteracy group has more children than the literacy group.

```{r, fig.pos="H"}
#| warning: false
#| echo: false
#| message: false
#| label: fig-rr
#| fig-cap: "Relative Risk"
possi_model_ci <- as.data.frame(exp(confint(possi_model)[-1,]))
possi_model_ci$level = gsub(paste(names(possi_model$xlevels), collapse='|'),
   "", rownames(possi_model_ci))
possi_model_ci$variable=unlist(strsplit(rownames(possi_model_ci),possi_model_ci$level))
possi_model_ci$x = 1:nrow(possi_model_ci)
possi_model_ci$cex = sqrt(1/apply(log(possi_model_ci[,1:2]),1,diff))
forXaxis = tapply(possi_model_ci$x, possi_model_ci$variable, mean)

matplot(possi_model_ci[,1:2], type='n',
   xaxt='n', bty='n', xaxt='n',
   log='y', ylab='RR')

segments(possi_model_ci$x, possi_model_ci[,1],
   possi_model_ci$x, possi_model_ci[,2])

 points(possi_model_ci$x,
   exp(possi_model$coef[-1]), pch=15,
   cex = possi_model_ci$cex, col = '#00000030')
 
mtext(possi_model_ci$level, 1, at=possi_model_ci$x,
   las=3, line=-1)

mtext(names(forXaxis), 1, at=forXaxis,
   line=-2)
abline(h=1, lty=3)
```

More precise values can be seen in @tbl-msum. Compared to the reference group (those who married at ages 22–25), the RRs for those who married at ages 15–18 and 18–20 are 1.079 and 1.064 respectively, with confidence intervals that do not cross 1, indicating that their number of children is significantly higher than that of the reference group. The RR for those who married at ages 0–15 is approximately 1.051; however, its wide confidence interval and relatively high p-value suggest that this effect is not statistically significant. For those who married at ages 20–22, the RR is only 1.030 with a confidence interval that includes 1, indicating a minimal effect on the number of children, while the RRs for the groups that married at ages 25–30 and 30 or above are very close to 1 and have p-values far greater than 0.05, showing no significant difference in fertility compared to the reference group. Moreover, the impact of literacy on the number of children is quite pronounced: the “literacy = no” group has an RR of 1.173, with a confidence interval of [1.120, 1.228] that does not include 1, and a p-value of 0.000, indicating that the number of children among the illiterate is significantly higher than that among the literate.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-msum
#| tbl-cap: "Summary of The Final Model"

logRateMat = cbind(est=possi_model$coef, confint(possi_model, level=0.95))
knitr::kable(cbind(
    summary(possi_model)$coef, 
    exp(logRateMat)), 
  digits=3)
```







